---
title: "Year 1 stats spin-off -- plots and models"
author: Dr Jens Roeser 
#date: Feb 15 2021 # Feb 22 2021
output: 
  revealjs::revealjs_presentation:
    theme: night
    highlight: tango
    incremental: false
    transition: concave
    background_transition: zoom
    widescreen: true

bibliography: ["references.bib"]
csl: apa.csl
link-citations: yes

---

```{r setup, include=FALSE}
library(citr)
library(tidyverse)
library(magrittr)
library(psyntur)
library(ggthemes)
library(kableExtra)
library(knitr)
library(readxl)
library(extrafont)
library(broom)
options("kableExtra.html.bsTable" = T)
knitr::opts_chunk$set(echo = FALSE,
                      comment=NA, 
                      warning = FALSE,
                      message =FALSE)

theme_set(theme_classic(base_size = 16) +
          theme(panel.grid = element_line(colour = "grey95"),
                panel.background = element_rect(fill = "transparent")))
```


# Learning outcomes 

<div style="font-size:1.25em;">
After completing this lecture, the workshop and your own reading you should be able to ... 
</div>



# Overview

- table with plots and associated models
- decision tree for models


# Data visualisation

## Histogram

## Scatterplot

## Tukey boxplot


# Data modelling 

## Correlation

```{r}
N <- 1000
rs <- c(-.999, -.75, -.25, 0, .25, .75, .999)

cors <- map(rs, ~faux::rnorm_multi(N, 2, 100, 50, ., varnames = c("x", "y")) %>%
      mutate(correlation = paste0("italic('r')~'='~",.x))) %>%
  bind_rows() %>%
  mutate(order = rep(1:length(rs), each = N),
         correlation = factor(correlation, levels = unique(correlation)[unique(order)], ordered = T))
  
ggplot(cors, aes(x = x, y = y)) +
  geom_point(size = .1, alpha = .5) +
  facet_wrap(~correlation, labeller = label_parsed) +
  theme_bw() 
```


## Linear regression

## Independent *t*-test

## Paired samples *t*-test

## One-way ANOVA

## Test statistics and estimates

- F, t, r, $\beta$ (beta), p, df


# Model assumptions and evaluation








# Epilogue {data-transition="fade-out"}

## Summary {data-transition="fade-in fade-out"}



## Exam-prep advice {data-transition="fade-in fade-out"}



## Learning outcomes {data-transition="fade-in fade-out"}




## Useful textbook resources  {data-transition="fade-in fade-out"}




## References  {data-transition="fade-in fade-out"}

<style>
    .reveal section p {
    display: inline-block;
    font-size: 0.7em;
    line-height: 1.2em;
    vertical-align: top;
    horizontal-align: left;
    margin: 0 0 0 0;
    line-spacing: .5em;
  }
</style>





