---
title: "Workshop 18: Model evaluation and violations"
author: "Jens Roeser"
date: "Compiled `r Sys.Date()`"
#output: 
#  bookdown::html_document2: default
#  pdf_document: default
#  ss: styles.css
#  fig_caption: yes
#  theme: flatly
#  toc: no
#  toc_depth: 1
#  spacing: double
#  indent: true
bibliography: ../slides/references.bib
csl: apa.csl
link-citations: yes
output:
  rmdformats::readthedown:
  self_contained: false
fig_width: 2
fig_height: 2
thumbnails: true
lightbox: true
gallery: false
highlight: "kate"
toc_depth: 1
use_bookdown: false

---
  
  
  ```{r, include=FALSE}
library(citr) # CRTL + Shift + R for citations
library(kableExtra)
library(rmdformats)
library(knitr)
library(tidyverse)
library(psyntur)
set.seed(123)
theme_set(theme_bw())
```

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(width=90)


# Fit lm regression and anova style

# Go through lecture example (continuous predict)

# Give instructions for example with discrete predictor)

# extracting residuals

# histogram of residuals: normality

# Skew and kurtosis

# log transformation

# plot residuals against fitted data / predictor: independent -> equality of variance / linearity


```


# Lecture review

  - Parametric models (*t*-test, ANOVA, linear regression) make assumptions about their input (the data_.
  - Assumptions related to the normal distribution
    - data must be continuous
    - observations must be independent and identically distributed 
  - The normal distribution is assumed because of the central limit theorem
  - Homogeneity of variance
  - Linearity: continuous predictors must be linear

# Learning outcomes 

This workshop, along with the lecture on model assumptions should give you the opportunity to be able to



For revision ....

# Setup

You need the package `psyntur`:

```{r}
library(psyntur) 
packageVersion("psyntur") # should be version 0.0.2
```


# Fit two models

## Model with categorical predictor

## Model with continuous predictor

# Residuals

# Normal distribution

## Continuous data

## Transformation

# Linearity

# Homogeneity of variance



# References

<div id="refs"></div>
  